\vspace{-2em} Hamiltonian Privilege

### 2021-04-26 Articles/2020-ClassicalMechanicsPhil/Submissions/BJPS/classical-phil-article-BJPS.tex
Update classical-phil-article-BJPS.tex
Cut most of section 6.2 (discussion of extended phase space formalism). Eliminated many footnotes and trimmed down other extraneous text (e.g. brief discussion of discrete case). Cut \geq 1185 words

% 300 words maximum for BJPS
\vspace{-3em}
\bs
\begin{abstract}
\noindent We argue that Hamiltonian mechanics is more fundamental than Lagrangian mechanics. Our argument provides a non-metaphysical strategy for privileging one formulation of a theory over another: a formulation is more fundamental at least when it is more general. We illustrate this criterion through a novel interpretation of classical mechanics in terms of three physical conditions. Two of these conditions suffice for recovering Hamiltonian mechanics. A third condition is necessary for recovering Lagrangian mechanics. This shows that Lagrangian mechanics is a proper subset of Hamiltonian mechanics, elucidating the connection between these two formalisms. Using our framework, we argue that the invariant phase space density is physically fundamental. Since Hamiltonian mechanics provides the proper setting for this density, we gain another reason for privileging it over Lagrangian mechanics. Finally, we interpret Lagrangian mechanics within the Hamiltonian framework, providing a geometric interpretation of the principle of stationary action. We rebut arguments for privileging Lagrangian mechanics, showing where they go wrong. 
\end{abstract}
\es

@@

In the context of comparing Hamiltonian and Lagrangian mechanics, three interpretive stances have emerged. \textcites[]{North} has argued for privileging Hamiltonian mechanics, on the grounds that it requires less mathematical structure to formulate. In contrast, \textcites[]{Curiel} recommends privileging Lagrangian mechanics, arguing that it isolates physically-significant kinematical constraints that all classical systems must satisfy. Finally, we can interpret Lagrangian and Hamiltonian mechanics as theoretically equivalent in a precise mathematical sense. Within a subclass of classical mechanical systems known as the \textit{hyper-regular domain}, the two formalisms are categorically equivalent \parencites[]{Teh}{Barrett2}. Categorical equivalence provides a natural interpretive criterion for theoretical equivalence because it yields mutual inter-translatability between theory formulations. This equivalence motivates the view that there are no physically significant differences between these two formalisms, at least within the hyper-regular domain.

@@

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we examine the gauge redundancy of both the conjugate momentum (Section~\ref{gauge_momentum}) and the Lagrangian (Section~\ref{gauge}).

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by `fundamental.' Of the various notions of fundamentality, we will mean something entirely non-mysterious: a formalism is more fundamental provided that it is logically more general. In this case, the Lagrangian framework is less general because it requires that an additional condition hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.\footnote{One might worry that our criterion for fundamentality is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where `aboutness' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality or essence must address. Hence, it is not a special problem facing our criterion.}

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space.\footnote{Since photons travel at constant speed, we can recover the direction of momentum---but not the magnitude---from the trajectory. Two photons with different momenta can follow the same trajectory in physical space.} Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.

Of course, different definitions of `Hamiltonian' and `Lagrangian systems' can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they constitute the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications. For details of the Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as neither Lagrangian nor Hamiltonian. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics.\footnote{For discussion of the numerous conceptual problems that arise within different approaches to classical mechanics (including point-masses, rigid-bodies, and continua) see \textcites{Wilson}. In Wilson's terminology, the framework we apply here may be yet another `theory facade' for classical physics, but a fruitful one nonetheless.} This supports our contention that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by `fundamental.'

@@

As indicated at the start of Section~\ref{privileging}, we use `fundamentality' in a non-metaphysical fashion. Some differences in fundamentality are differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of `fundamental,' discussed in Section~\ref{density}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of `structure.' Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Owing to our epistemic inabilities to access these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant densities (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. Precisely this scenario occurs in the context of Lagrangian mechanics, which requires the additional condition of motion equivalence to hold. Whereas North views Lagrangian mechanics as containing ``excess, superfluous structure'' \parencites*[75]{North}, we take this additional condition as optional but still physically significant. Thus, while it is illuminating to identify the minimum mathematical structures necessary for a formulation, additional structure is not automatically rendered physically insignificant.

Additionally, whereas North \parencites*[76]{North} seeks to determine the `real' fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure ontologically, but as a formal device for encoding physical information or a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental in a classical world does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically and conceptually more fundamental. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics.\footnote{These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we typically use below.} Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics further supports viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry'' \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some `unphysical' coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling.

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids the criticisms that Swanson and Halvorson develop against North. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the assumption of motion equivalence. The assumption of motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are physically fundamental and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.

@@

\begin{quotation}
\bs \noindent
\textit{Infinitesimal reducibility}: the state of the system is reducible to the state of its infinitesimal parts. That is, giving the state of the whole system is equivalent to giving the state of its parts, which is equivalent to giving the states of those parts, and so on. \es
\end{quotation}

@@

If $\mathcal{S}$ is a discrete space with $U \subseteq \mathcal{S}$, we define $\mu(U) = \sum\limits_{s \in U} \rho_c(s)$ so that $\mu(\mathcal{S})$ gives us the size of the system and $\mu(U)/\mu(\mathcal{S})$ gives us the fraction of the particles that can be found in the region $U$.\footnote{The discrete case is far less interesting because discrete quantities cannot change continuously over time, meaning that they must be constant. We discuss this elsewhere.}
% % % For details see \textcites[]{AoPPhy1}. [omit this for review]}
% % XXX could try to find justification/derivation for the state space being a manifold in the Kevin Kelly book, or could cite the math paper with Greenfield; proposition 6.2. Try to include a better gloss/motivation for this claim (basic idea: if time is continuous, trajectories in state space must be continuous, local structure must be compatible). Curiel paper might also have a brief justification/motivation, if I recall

In the continuous case, $\mathcal{S}$ is a manifold, and we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. We will call a set of quantities $\xi^a$ \textit{state variables} provided that they fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates' and $\xi$ is a coordinate map. We will have multiple manifolds, so to avoid confusion, we will call the coordinates of the state space `state variables'. As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} In other words, $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument.

Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$ (i.e. the choice of state variable map $\xi$). Hence, the density $\rho_{c, \xi}$ transforms as a scalar under state variable changes.\footnote{For two state variable maps $\xi$ and $\hat{\xi}$, we have $\rho_c(s) = (\rho_c \circ \xi^{-1} \circ \xi)(s) = (\rho_c \circ \hat{\xi}^{-1} \circ \hat{\xi})(s)$, which entails that $\rho_c(s) = \rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Here, $\xi(s) \equiv \xi^a$ and $\hat{\xi}(s) \equiv \hat{\xi}^b$.} Likewise, the integral $\int_U \rho_c(s) d\mathcal{S} $ is independent of the choice of state variables, so $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$. By the change of variables formula, we also know that $\int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b= \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.\footnote{The change of variables transformation from $\xi^a$ to $\hat{\xi}^b$ is given by $\phi \equiv \hat{\xi} \circ \xi^{-1}$: $\mathbb{R}^n \to \mathbb{R}^n $. For convenience, we denote $\rho_{c, \hat{\xi}} (\phi(\xi^a))$ by $\rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Likewise, in the numerator of the Jacobian, we denote $\partial \phi$ by $\partial \hat{\xi}^b$, where $\hat{\xi}^b$ is here understood as a function of $\xi^a$.} Here, $ \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ is the determinant of the Jacobian for this change of state variables. From the preceding two equations, we see that $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$. Matching integrands, we have $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$. From the scalar transformation property, $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Hence, we see that $\rho_{c, \hat{\xi}} (\hat{\xi}^b) =  \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$, entailing that the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless. We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the classical phase space must be at least a symplectic manifold.

We define a \textit{unit variable} $q \in \xi^a$ as a particular state variable that defines a unit. For instance, $q$ might be position, defined in units of meters. Two unit variables are \textit{independent} if one unit variable can be changed without changing the other. For example, one can measure the vertical position in meters while measuring the horizontal position in miles.  In general, the unit variables $q^i$ are the subset of the state variables that define the unit system, which is needed to relate the properties of the system to the measured values. The $q^i$ determine the units not only for themselves, but for all other state variables. A \textit{change of units} is therefore a map $\hat{q}^j = \hat{q}^j(q^i)$ that defines new unit variables in terms of the old unit variables alone. Since the unit variables define the units for all state variables, this change must induce a unique change of state variables $\hat{\xi}^b = \hat{\xi}^b(\xi^a)$ for all variables.

Let us consider the simplest case: suppose that a single unit variable $q $ suffices to define the unit system for the particle state space. Then, the state variables are given by a set $\xi^a = \{ q, k^\alpha \}$, where \textit{prima facie} the index $\alpha$ could be any non-negative integer. We will show that, in fact, there must be exactly one $k^\alpha$ variable, conjugate to the single $q$ variable.

>>>>

% 300 words maximum for BJPS
\vspace{-3em}
\bs
\begin{abstract}
\noindent We argue that Hamiltonian mechanics is more fundamental than Lagrangian mechanics. Our argument provides a non-metaphysical strategy for privileging one formulation of a theory over another: a formulation is more fundamental at least when it is more general. We illustrate this criterion through a novel interpretation of classical mechanics in terms of three physical conditions. Two of these conditions suffice for recovering Hamiltonian mechanics. A third condition is necessary for recovering Lagrangian mechanics. This shows that Lagrangian systems are a proper subset of Hamiltonian systems, elucidating the connection between these two formalisms. Using our framework, we argue that the invariant phase space density is physically fundamental. Since Hamiltonian mechanics provides the proper setting for this density, we gain another reason for privileging it over Lagrangian mechanics. We interpret Lagrangian mechanics within the Hamiltonian framework, providing a geometric interpretation of the principle of stationary action. Finally, we rebut arguments for privileging Lagrangian mechanics, showing where they go wrong. 
\end{abstract}
\es

@@

In the context of comparing Hamiltonian and Lagrangian mechanics, three interpretive stances have emerged. \textcites[]{North} has argued for privileging Hamiltonian mechanics on the grounds that it requires less mathematical structure to formulate. In contrast, \textcites[]{Curiel} recommends privileging Lagrangian mechanics, arguing that it isolates physically-significant kinematical constraints that all classical systems must satisfy. Finally, we can interpret Lagrangian and Hamiltonian mechanics as theoretically equivalent in a precise mathematical sense. Within a subclass of systems known as the \textit{hyper-regular domain}, the two formalisms are categorically equivalent \parencites[]{Teh}{Barrett2}. Categorical equivalence provides a natural interpretive criterion for theoretical equivalence because it yields mutual inter-translatability between theory formulations. This equivalence motivates the view that there are no physically significant differences between these two formalisms, at least within the hyper-regular domain.

@@

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. 
%Along the way, we examine the gauge redundancy of both the conjugate momentum (Section~\ref{gauge_momentum}) and the Lagrangian (Section~\ref{gauge}).

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by `fundamental.' Of the various notions of fundamentality, we will mean something entirely non-mysterious: a formalism is more fundamental provided that it is logically more general. In this case, the Lagrangian framework is less general because it requires an additional condition to hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.\footnote{One might worry that our criterion for fundamentality is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where `aboutness' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality or essence must address. Hence, it is not a special problem facing our criterion.}

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space.\footnote{Since photons travel at constant speed, we can recover the direction of momentum---but not the magnitude---from the trajectory. Two photons with different momenta can follow the same trajectory in physical space.} Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.

Of course, different definitions of `Hamiltonian' and `Lagrangian' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they constitute the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications. For details of the Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the derivational details of our argument in Section~\ref{derivation}, we describe how we avoid central problems facing North's privileging of Hamiltonian mechanics. 
%We view non-dissipative systems as neither Lagrangian nor Hamiltonian. 
%\footnote{For discussion of the numerous conceptual problems that arise within different approaches to classical mechanics (including point-masses, rigid-bodies, and continua) see \textcites{Wilson}. In Wilson's terminology, the framework we apply here may be yet another `theory facade' for classical physics, but a fruitful one nonetheless.}

@@

As indicated at the start of Section~\ref{privileging}, we use `fundamentality' in a non-metaphysical fashion. Some differences in fundamentality are differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of `fundamental,' discussed in Section~\ref{density}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of `structure.' Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which scientific anti-realists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Owing to the epistemic difficulty of accessing these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant densities (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. Precisely this scenario occurs in the context of Lagrangian mechanics, which requires the additional condition of motion equivalence to hold. Whereas North views Lagrangian mechanics as containing ``excess, superfluous structure'' \parencites*[75]{North}, we take this additional condition as optional but still physically significant. Thus, while it is illuminating to identify the minimum mathematical structures necessary for a formulation, additional structure is not automatically rendered physically insignificant.

Additionally, whereas North \parencites*[76]{North} seeks to determine the `real' fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure ontologically, but as a formal device for encoding physical information or a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental in a classical world does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically and conceptually more fundamental. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics. These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we recover in Section~\ref{deterministic}.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics further supports viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry'' \parencites*[46]{Teh}.}

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids these problems. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the motion equivalence condition. Motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are physically fundamental and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.

@@

\begin{quotation}
\bs \noindent
\textit{Infinitesimal reducibility}: the state of the system is reducible to the state of its infinitesimal parts. That is, specifying the state of the whole system is equivalent to specifying the state of its parts.
\es
\end{quotation}

@@

%If $\mathcal{S}$ is a discrete space with $U \subseteq \mathcal{S}$, we define $\mu(U) = \sum\limits_{s \in U} \rho_c(s)$ so that $\mu(\mathcal{S})$ gives us the size of the system and $\mu(U)/\mu(\mathcal{S})$ gives us the fraction of the particles that can be found in the region $U$.\footnote{The discrete case is far less interesting because discrete quantities cannot change continuously over time, meaning that they must be constant. We discuss this elsewhere.}
% % % For details see \textcites[]{AoPPhy1}. [omit this for review]}
% % XXX could try to find justification/derivation for the state space being a manifold in the Kevin Kelly book, or could cite the math paper with Greenfield; proposition 6.2. Try to include a better gloss/motivation for this claim (basic idea: if time is continuous, trajectories in state space must be continuous, local structure must be compatible). Curiel paper might also have a brief justification/motivation, if I recall

When physical quantities change continuously over time, $\mathcal{S}$ is a manifold, and we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. State variables $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates' and $\xi$ is a coordinate map. We will have multiple manifolds, so to avoid confusion, we will call the coordinates of the state space `state variables'. As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument.

Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$ (i.e. the choice of state variable map $\xi$). Hence, the density $\rho_{c, \xi}$ transforms as a scalar under state variable changes.\footnote{For two state variable maps $\xi$ and $\hat{\xi}$, we have $\rho_c(s) = (\rho_c \circ \xi^{-1} \circ \xi)(s) = (\rho_c \circ \hat{\xi}^{-1} \circ \hat{\xi})(s)$, which entails that $\rho_c(s) = \rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Here, $\xi(s) \equiv \xi^a$ and $\hat{\xi}(s) \equiv \hat{\xi}^b$.} Likewise, the integral $\int_U \rho_c(s) d\mathcal{S} $ is independent of the choice of state variables, so $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$. By the change of variables formula, we also know that $\int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b= \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.\footnote{The change of variables transformation from $\xi^a$ to $\hat{\xi}^b$ is given by $\phi \equiv \hat{\xi} \circ \xi^{-1}$: $\mathbb{R}^n \to \mathbb{R}^n $. For convenience, we denote $\rho_{c, \hat{\xi}} (\phi(\xi^a))$ by $\rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Likewise, in the numerator of the Jacobian, we denote $\partial \phi$ by $\partial \hat{\xi}^b$, where $\hat{\xi}^b$ is here understood as a function of $\xi^a$.} Here, $ \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ is the determinant of the Jacobian for this change of state variables. Matching integrands from the preceding two equations, we have $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$. From the scalar transformation property, $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Hence, we see that $\rho_{c, \hat{\xi}} (\hat{\xi}^b) =  \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$, entailing that the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless. We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the classical phase space must be at least a symplectic manifold. 
%From the preceding two equations, we see that $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.

We define a \textit{unit variable} $q \in \xi^a$ as a particular state variable that defines a unit. For instance, $q$ might be position, defined in units of meters. Two unit variables are \textit{independent} if one unit variable can be changed without changing the other. For example, one can measure vertical position in meters while measuring horizontal position in miles.  In general, the unit variables $q^i$ are the subset of the state variables that define the unit system, which relates the properties of the system to measured values. The $q^i$ determine the units not only for themselves, but for all other state variables. A \textit{change of units} is therefore a map $\hat{q}^j = \hat{q}^j(q^i)$ that defines new unit variables in terms of old unit variables. Since the unit variables define the units for all state variables, this change must induce a unique change of state variables $\hat{\xi}^b = \hat{\xi}^b(\xi^a)$ for all variables.

Let us consider the simplest case: suppose that a single unit variable $q $ suffices to define the unit system for the particle state space. Then, the state variables are a set $\xi^a = \{ q, k^\alpha \}$, where \textit{prima facie} the index $\alpha$ could be any non-negative integer. We will show that, in fact, there must be exactly one $k^\alpha$ variable, conjugate to the single $q$ variable.
